---
title: 'MAM 2022 Pre-programme Assignment'
author: "Yifan Yang"
date: "28/08/2021"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    toc: yes
    toc_float: yes
---

```{r load-libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(gapminder)  # gapminder dataset
library(here)
library(janitor)
```

The goal is to test your software installation, to demonstrate competency in Markdown, and in the basics of `ggplot`.

# Task 1: Short biography written using markdown


## Biography of Esther

__Hi everyone!__ My name is _Esther_. I am 22 years old. I was born in 1999, and [here](https://www.historic-newspapers.co.uk/blog/1999-events/) is what happened in that year. It is sad that this website does not include my my birth as one of the events.

During my free time, I like to

* travel

* cook

* and listen to music.


# Task 2: `gapminder` country comparison

You have seen the `gapminder` dataset that has data on life expectancy, population, and GDP per capita for 142 countries from 1952 to 2007. To get a glimpse of the dataframe, namely to see the variable names, variable types, etc., we use the `glimpse` function. We also want to have a look at the first 20 rows of data.

```{r}
glimpse(gapminder)

head(gapminder, 20) # look at the first 20 rows of the dataframe

```

Your task is to produce two graphs of how life expectancy has changed over the years for the `country` and the `continent` you come from.

I have created the `country_data` and `continent_data` with the code below.

```{r}
country_data <- gapminder %>% 
            filter(country == "China") # just choosing Greece, as this is where I come from

continent_data <- gapminder %>% 
            filter(continent == "Asia")
```

First, create a plot of life expectancy over time for the single country you chose. Map `year` on the x-axis, and `lifeExp` on the y-axis. You should also use `geom_point()` to see the actual data points and `geom_smooth(se = FALSE)` to plot the underlying trendlines. You need to remove the comments **\#** from the lines below for your code to run.

```{r, lifeExp_one_country,message=FALSE}
plot1 <- ggplot(data = country_data, mapping = aes(x = year, y = lifeExp))+
geom_point() +
geom_smooth(se = FALSE)
#  NULL 

plot1
```

Next we need to add a title. Create a new plot, or extend plot1, using the `labs()` function to add an informative title to the plot.

```{r, lifeExp_one_country_with_label, message=FALSE}
plot1<- plot1 +
labs(title = " Life expectancy over time for China",
x = "Year",
y = "Life expectancy") 
#   NULL


plot1
```

Secondly, produce a plot for all countries in the *continent* you come from. (Hint: map the `country` variable to the colour aesthetic. You also want to map `country` to the `group` aesthetic, so all points for each country are grouped together).

```{r lifeExp_one_continent,message=FALSE}
ggplot(data = continent_data, mapping = aes(x = year , y = lifeExp , colour= country, group = country ))+
geom_point() + 
geom_smooth(se = FALSE) 
#   NULL
```

Finally, using the original `gapminder` data, produce a life expectancy over time graph, grouped (or faceted) by continent. We will remove all legends, adding the `theme(legend.position="none")` in the end of our ggplot.

```{r lifeExp_facet_by_continent,message=FALSE}
ggplot(data = gapminder , mapping = aes(x = year , y = lifeExp , colour= country, group = continent))+
geom_point() + 
geom_smooth(se = FALSE) +
facet_wrap(~continent) +
theme(legend.position="none") 
#   NULL
```

Given these trends, what can you say about life expectancy since 1952? Again, don't just say what's happening in the graph. Tell some sort of story and speculate about the differences in the patterns.

> Type your answer after this blockquote.

The life expectancy for all continents has shown an increasing trend, with that for Africa gradually flattening  after 1985. While Oceania and Europe have the highest average life expectancy, Africa ranks the last, with Asia and Americas above it. Generally, the life expectancy can be seen as positively related to the continent's economic growth.  

# Task 3: Brexit vote analysis

We will have a look at the results of the 2016 Brexit vote in the UK. First we read the data using `read_csv()` and have a quick glimpse at the data

```{r load_brexit_data, warning=FALSE, message=FALSE}
brexit_results <- read_csv(here::here("data","brexit_results.csv"))


glimpse(brexit_results)
```

The data comes from [Elliott Morris](https://www.thecrosstab.com/), who cleaned it and made it available through his [DataCamp class on analysing election and polling data in R](https://www.datacamp.com/courses/analyzing-election-and-polling-data-in-r).

Our main outcome variable (or y) is `leave_share`, which is the percent of votes cast in favour of Brexit, or leaving the EU. Each row is a UK [parliament constituency](https://en.wikipedia.org/wiki/United_Kingdom_Parliament_constituencies).

To get a sense of the spread, or distribution, of the data, we can plot a histogram, a density plot, and the empirical cumulative distribution function of the leave % in all constituencies.

```{r brexit_histogram, warning=FALSE, message=FALSE}

# histogram
ggplot(brexit_results, aes(x = leave_share)) +
  geom_histogram(binwidth = 2.5)+
  labs(title='Vote outcome',x='Leave_share',y='Number of constituencies',subtitle='Histogram of the number count of constituencies vs. their corresponding percentage of votes')

# density plot-- think smoothed histogram
ggplot(brexit_results, aes(x = leave_share)) +
  geom_density()+
  labs(title='Vote outcome',x='Leave_share',y='Density',subtitle='Density plot of the number count of constituencies vs. their corresponding percentage of votes')


# The empirical cumulative distribution function (ECDF) 
ggplot(brexit_results, aes(x = leave_share)) +
  stat_ecdf(geom = "step", pad = FALSE) +
  scale_y_continuous(labels = scales::percent)+
  labs(title='Vote outcome',x='Leave_share',y='Cumulative percentage',subtitle='Cumulative distribution function plot of the number count of constituencies vs. their corresponding percentage of votes')
  


```

One common explanation for the Brexit outcome was fear of immigration and opposition to the EU's more open border policy. We can check the relationship (or correlation) between the proportion of native born residents (`born_in_uk`) in a constituency and its `leave_share`. To do this, let us get the correlation between the two variables

```{r brexit_immigration_correlation}
brexit_results %>% 
  select(leave_share, born_in_uk) %>% 
  cor()
```

The correlation is almost 0.5, which shows that the two variables are positively correlated.

We can also create a scatterplot between these two variables using `geom_point`. We also add the best fit line, using `geom_smooth(method = "lm")`.

```{r brexit_immigration_plot,message=FALSE}
ggplot(brexit_results, aes(x = born_in_uk, y = leave_share)) +
  geom_point(alpha=0.3) +
  
  # add a smoothing line, and use method="lm" to get the best straight-line
  geom_smooth(method = "lm") + 
  
  # use a white background and frame the plot with a black box
  theme_bw() +
  
  labs(title='Vote outcome',x='Proportion of native born residents',y='Leave_share',subtitle='Scatter plot of the leave share vs. the proportion of native born residents within this leave share')
  NULL
```

You have the code for the plots, I would like you to revisit all of them and use the `labs()` function to add an informative title, subtitle, and axes titles to all plots.

What can you say about the relationship shown above? Again, don't just say what's happening in the graph. Tell some sort of story and speculate about the differences in the patterns.

> Type your answer after, and outside, this blockquote.

# Task 4: Animal rescue incidents attended by the London Fire Brigade

[The London Fire Brigade](https://data.london.gov.uk/dataset/animal-rescue-incidents-attended-by-lfb) attends a range of non-fire incidents (which we call 'special services'). These 'special services' include assistance to animals that may be trapped or in distress. The data is provided from January 2009 and is updated monthly. A range of information is supplied for each incident including some location information (postcode, borough, ward), as well as the data/time of the incidents. We do not routinely record data about animal deaths or injuries.

Please note that any cost included is a notional cost calculated based on the length of time rounded up to the nearest hour spent by Pump, Aerial and FRU appliances at the incident and charged at the current Brigade hourly rate.

```{r load_animal_rescue_data, warning=FALSE, message=FALSE}

url <- "https://data.london.gov.uk/download/animal-rescue-incidents-attended-by-lfb/8a7d91c2-9aec-4bde-937a-3998f4717cd8/Animal%20Rescue%20incidents%20attended%20by%20LFB%20from%20Jan%202009.csv"

animal_rescue <- read_csv(url,
                          locale = locale(encoding = "CP1252")) %>% 
  janitor::clean_names()


glimpse(animal_rescue)
```
One of the more useful things one can do with any data set is quick counts, namely to see how many observations fall within one category. For instance, if we wanted to count the number of incidents by year, we would either use `group_by()... summarise()` or, simply [`count()`](https://dplyr.tidyverse.org/reference/count.html)

```{r, instances_by_calendar_year}

animal_rescue %>% 
  dplyr::group_by(cal_year) %>% 
  summarise(count=n())

animal_rescue %>% 
  count(cal_year, name="count")

```

Let us try to see how many incidents we have by animal group. Again, we can do this either using group_by() and summarise(), or by using count()

```{r, animal_group_percentages}
animal_rescue %>% 
  group_by(animal_group_parent) %>% 
  
  #group_by and summarise will produce a new column with the count in each animal group
  summarise(count = n()) %>% 
  
  # mutate adds a new column; here we calculate the percentage
  mutate(percent = round(100*count/sum(count),2)) %>% 
  
  # arrange() sorts the data by percent. Since the default sorting is min to max and we would like to see it sorted
  # in descending order (max to min), we use arrange(desc()) 
  arrange(desc(percent))


animal_rescue %>% 
  
  #count does the same thing as group_by and summarise
  # name = "count" will call the column with the counts "count" ( exciting, I know)
  # and 'sort=TRUE' will sort them from max to min
  count(animal_group_parent, name="count", sort=TRUE) %>% 
  mutate(percent = round(100*count/sum(count),2))


```

Do you see anything strange in these tables? 

* Most animals rescued fall into the group of Cat, Bird and Dog, while very few Squirrel and Unknown - Heavy Livestock Animal cases were present. Incidents related to Squirrel and Unknown - Heavy Livestock Animal may be more difficult to be spotted or reported compared to those common pet animal groups.

Finally, let us have a loot at the notional cost for rescuing each of these animals. As the LFB says,

> Please note that any cost included is a notional cost calculated based on the length of time rounded up to the nearest hour spent by Pump, Aerial and FRU appliances at the incident and charged at the current Brigade hourly rate.

There is two things we will do:

1. Calculate the mean and median `incident_notional_cost` for each `animal_group_parent`
2. Plot a boxplot to get a feel for the distribution of `incident_notional_cost` by `animal_group_parent`.


Before we go on, however, we need to fix `incident_notional_cost` as it is stored as a `chr`, or character, rather than a number.

```{r, parse_incident_cost,message=FALSE, warning=FALSE}

# what type is variable incident_notional_cost from dataframe `animal_rescue`
typeof(animal_rescue$incident_notional_cost)

# readr::parse_number() will convert any numerical values stored as characters into numbers
animal_rescue <- animal_rescue %>% 

  # we use mutate() to use the parse_number() function and overwrite the same variable
  mutate(incident_notional_cost = parse_number(incident_notional_cost))

# incident_notional_cost from dataframe `animal_rescue` is now 'double' or numeric
typeof(animal_rescue$incident_notional_cost)

```

Now that incident_notional_cost is numeric, let us quickly calculate summary statistics for each animal group. 


```{r, stats_on_incident_cost,message=FALSE, warning=FALSE}

animal_rescue %>% 
  
  # group by animal_group_parent
  group_by(animal_group_parent) %>% 
  
  # filter resulting data, so each group has at least 6 observations
  filter(n()>6) %>% 
  
  # summarise() will collapse all values into 3 values: the mean, median, and count  
  # we use na.rm=TRUE to make sure we remove any NAs, or cases where we do not have the incident cos
  summarise(mean_incident_cost = mean (incident_notional_cost, na.rm=TRUE),
            median_incident_cost = median (incident_notional_cost, na.rm=TRUE),
            sd_incident_cost = sd (incident_notional_cost, na.rm=TRUE),
            min_incident_cost = min (incident_notional_cost, na.rm=TRUE),
            max_incident_cost = max (incident_notional_cost, na.rm=TRUE),
            count = n()) %>% 
  
  # sort the resulting data in descending order. You choose whether to sort by count or mean cost.
  arrange(desc(count))

```



Compare the mean and the median for each animal group. what do you think this is telling us?

* The median values are all smaller than the mean values except for Squirrels, Rabbit and Ferret. This means that for these species, there were lots of expensive rescue cases which have driving the mean cost higher than the median.

Anything else that stands out? Any outliers?

* For the three Unknown groups, Cow and Deer, their median values are much lower than their mean values. The minimum incident cost for dogs, which is 0, could be an outlier. Besides, for animal groups of a high incident count (such as Cat, Bird and Dog), their maximum incident costs are probably also outliers, as they are relatively too high for the mean and median values. If these expensive rescues were common, the mean value should be much higher than the median value.

Finally, let us plot a few plots that show the distribution of incident_cost for each animal group.

```{r, plots_on_incident_cost_by_animal_group,message=FALSE, warning=FALSE}

# base_plot
base_plot <- animal_rescue %>% 
  group_by(animal_group_parent) %>% 
  filter(n()>6) %>% 
  ggplot(aes(x=incident_notional_cost))+
  facet_wrap(~animal_group_parent, scales = "free")+
  theme_bw()

base_plot + geom_histogram()
base_plot + geom_density()
base_plot + geom_boxplot()
base_plot + stat_ecdf(geom = "step", pad = FALSE) +
  scale_y_continuous(labels = scales::percent)



```

Which of these four graphs do you think best communicates the variability of the `incident_notional_cost` values? Also, can you please tell some sort of story (which animals are more expensive to rescue than others, the spread of values) and speculate about the differences in the patterns.

* The first set of diagrams (histogram) best communicates the above information. We can infer that for Bird, Cat, Deer, Dog, Fox, Hamster, Horse, Squirrel, Unknown - Domestic Animal Or Pet, Unknown - Heavy Livestock Animal and Unknown - Wild Animal, there costs are relatively more concentrated in a certain range compared to the other groups. For those groups, the cost can vary a lot from case to case, which probably indicates that these animals live in a complex of environments and may have more unexpected incident. 

* Generally, Deer and Unknown - Wild Animal have their costs spread in a broader range than the other groups. While Horse are averagely more expensive to rescue, Cat, Deer, Fox, Cow, Bird, Dog and the three Unknown groups can be expensive in rare cases.

# Submit the assignment

Knit the completed R Markdown file as an HTML document (use the "Knit" button at the top of the script editor window) and upload it to Canvas.

## Details

If you want to, please answer the following

-   Who did you collaborate with: Zhicong Hu
-   Approximately how much time did you spend on this problem set: 3 hours
-   What, if anything, gave you the most trouble: :)
